---
title: "Calibrated specification curve analysis with speccal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrated specification curve analysis with speccal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, dpi = 96)
Sys.setenv(OMP_NUM_THREADS = "1")
```

A specification curve (multiverse) analysis fits every analysis a reasonable investigator could have run
and reports the distribution of estimates. Existing tools stop there. `speccal` adds the inference:
which specifications are defensible, whether the share of significant specifications exceeds what
confounding alone produces, how many of the specifications are effectively independent, whether the
analytic choice that moves the estimates most does so even when no association exists, and, for two
common special cases, whether a composite index is a sound way to enter its constituents and whether a
reported threshold is located by the data.

This vignette walks through the workflow on the bundled dataset `nhanes_perio`: 10,108 adults from
NHANES 2009-2014 with eleven periodontitis case definitions, ten blood-derived inflammatory markers
and their constituents, and 16 covariates. The accompanying paper analyses a grid of 15,840
specifications; to keep this document quick to build we use a smaller grid and fewer permutations. The
scripts in `inst/reproduce/` run the full analysis and check every number against the paper's ledger.

## 1. Data and survey design

```{r data}
library(speccal)
library(survey)
data(nhanes_perio)
options(survey.lonely.psu = "adjust")
des <- svydesign(id = ~SDMVPSU, strata = ~SDMVSTRA, weights = ~WT_TOTAL, data = nhanes_perio, nest = TRUE)
degf(des)
```

Everything that follows is design-based: weights, strata and primary sampling units enter every model,
and restricted samples are handled as domains (`subset()` on the design), never by filtering the data
before the design is built.

## 2. Define the axes

An axis is one analytic choice with two or more levels. Here: four case definitions, four nested
covariate sets (crude, demographics, plus socio-behavioural factors, plus clinical factors), two exposure
parameterisations and two sample restrictions, all design-based.

```{r axes, code = readLines("axes.R")}
```

```{r axes-print}
ax
```

`efp2` is the 2018 EFP/AAP surveillance adaptation at a 2 mm interdental threshold, which classifies 96%
of adults as cases. It is included deliberately: the point of the next steps is to let the data, not
the analyst, decide whether such a level belongs in the primary analysis.

## 3. Fit the grid

Three exposures: the white blood cell count, the neutrophil-to-lymphocyte ratio and the
albumin-to-globulin ratio.

The two steps that refit many models (the grid and the permutation null) take several minutes; their
results ship with the package so that this document builds quickly. The code is shown as it was run
(`data-raw/vignette_objects.R`).

```{r fit, eval = FALSE}
g <- spec_fit(ax, nhanes_perio, design = des, exposure = c("WBC", "NLR", "AGR"), cores = 8)
```

```{r load, include = FALSE}
obj <- readRDS(system.file("extdata", "vignette_objects.rds", package = "speccal"))
g <- obj$g; nul <- obj$nul
```

```{r fit-print}
g
head(as.data.frame(g)[, c("exposure", "outcome", "covset", "coding", "sample", "or", "p", "rr", "p_rr", "sep_flag")])
```

Every fit carries diagnostics: convergence, iteration count, the largest absolute coefficient and
standard error, and a quasi-separation flag. Marginal risk ratios and risk differences are computed by
standardisation, so conclusions can be checked on collapsible scales.

```{r curve, fig.height = 3.2}
plot(g)
```

## 4. A null distribution that keeps confounding intact

The usual permutation test shuffles the exposure across participants. That also destroys the
association between the exposure and the covariates, so it tests a world in which confounding does not
exist and is anticonservative whenever the exposure is confounded. The Freedman-Lane scheme regresses
the (log) exposure on the covariates, keeps the fitted part and permutes only the residuals within
strata. The null is then "no association given the covariates".

```{r null, eval = FALSE}
nul <- spec_null(g, scheme = "fl", B = 100, stratum = "SDMVSTRA", covariates = ax$covariates$full,
                 seed = 1, cores = 8)
```

```{r null-print}
covs_full <- ax$covariates$full
nul
```

One hundred permutations are enough to illustrate the machinery; the paper uses 300, and levels whose
false-positive rate lies close to the threshold need that many to be classified reliably. Every replicate is stored
specification by specification, which is what the calibration below needs.

## 5. Let the null define which specifications are defensible

For every level of every axis, the false-positive rate under the null is the share of null
specifications with P < 0.05. The threshold (here 10%, twice the nominal level) is set before looking at
the observed results. Case definitions and codings are assessed within the two adequately adjusted
covariate sets so that their behaviour is separated from residual confounding. The `within` levels must themselves
be defensible; `spec_defensible()` warns if they are not, because assessing a definition inside a
confounded model inflates its false-positive rate.

```{r defensible}
def <- spec_defensible(nul, threshold = 0.10, within = list(covset = c("socio", "full")))
def
plot(def)
keep <- spec_keep(def, g)
table(keep)
table(spec_keep(def, g, threshold = 0.15))   # sensitivity threshold
```

Not every classification is clear-cut. In this small grid the full-sample level has a false-positive rate
just above 10% with an interval that includes the threshold, so it is excluded at 10% and retained at 15%.
Borderline levels are exactly why the threshold is fixed in advance and the results are reported at the
sensitivity thresholds as well.

The crude and demographics-only models produce false positives far above the nominal rate and are
excluded, as is the near-saturated 2 mm definition, whose models inherit the instability of an outcome
with almost no non-cases. The interval on each rate
comes from the variation across permutations, because the specifications within one permutation are
correlated and a binomial interval would be too narrow. The criterion cannot detect over-adjustment: a
null of no association produces no false positives when a collider is adjusted. Pair it with a causal
argument for the covariates that are contested.

## 6. Calibrated summary: how much of "80% significant" is real?

```{r summary}
sm <- spec_summary(g, nul, keep = keep)
sm
plot(sm, null = nul, keep = keep, grid = g)
```

Three things are reported for each exposure that the raw share does not give:

* `p_perm`: the probability, under the null, of a share of significant specifications in the modal
  direction at least as large as the one observed.
* `n_eff_var` and `n_eff_liji`: the effective number of independent specifications, from the variance of
  the null share and from the eigenvalues of the correlation matrix of the estimates. Specifications of
  one exposure share the same participants, so they behave like far fewer independent tests.
* `ci_lo`, `ci_hi`: an approximate interval for the observed share using the effective number in place
  of the nominal denominator. The permutation P value remains the formal test.

## 7. Which analytic choice moves the estimate, and would it do so under the null?

```{r decompose, fig.height = 6}
vd <- spec_decompose(g, nul, interactions = TRUE, equal_cardinality = list(axis = "outcome", k = 2))
vd
plot(vd)
```

The decomposition is descriptive (the estimates are deterministic functions of one dataset), so no F
tests are attached. Three safeguards are built in: pairwise interactions show how much of the residual
is structured; the equal-cardinality draws show whether an axis dominates only because it has more
levels; and the same decomposition under the null shows how large a share an axis attains when no
association exists. A share that is as large under the null as in the data describes how that axis
transmits confounding and sampling variability, not evidence about the association.

## 8. Does a composite index impose a structure the data reject?

On the log scale, NLR = log(neutrophils) - log(lymphocytes): entering NLR equals entering the two counts
with coefficients constrained to +1 and -1 times a common slope. The constraint is testable.

```{r constraint}
ax_ct <- spec_axes(outcome = ax$outcome, covariates = ax$covariates, coding = "log2", sample = ax$subsets$sample)
ct <- composite_constraint(nhanes_perio, des, composite = "NLR", constituents = c(NEU = 1, LYM = -1), axes = ax_ct)
ct
summary(ct, exclude = list(outcome = "efp2"))
composite_coefficients(nhanes_perio, des, "NLR", c(NEU = 1, LYM = -1), outcome = "y_cdc_any", covariates = covs_full)
```

The neutrophil count carries the association and the lymphocyte count does not, whereas the ratio forces
them to enter with equal weight and opposite sign. Compare with the albumin-to-globulin ratio, whose
constraint the data accept:

```{r constraint-agr}
ct_agr <- composite_constraint(nhanes_perio, des, "AGR", c(ALB = 1, GLOB = -1), axes = ax_ct)
ct_agr
composite_coefficients(nhanes_perio, des, "AGR", c(ALB = 1, GLOB = -1), outcome = "y_cdc_any", covariates = covs_full)
```

## 9. Is a threshold located by the data or by the search range?

A two-segment model is fitted on the log2 exposure with the breakpoint chosen by minimum deviance over a
grid of candidates at fixed percentiles. The bootstrap distribution of the breakpoint, and in particular
the share of replicates that select the first or last candidate, shows whether the interval is set by
the data. The Rao-Wu rescaled bootstrap respects the survey design; the naive scheme is provided for
comparison and understates the variance when strata contain few units.

```{r breakpoint, fig.height = 3.5}
bp <- breakpoint_boot(nhanes_perio, des, exposure = "AGR", outcome = "y_cdc_any", covariates = covs_full,
                      grid = c(0.05, 0.95, 21), B = 100, resampling = "rao-wu", seed = 2,
                      sensitivity_grid = c(0.15, 0.85, 13))
bp
plot(bp)
```

## 10. Reading the output together

The workflow separates four claims that are usually reported as one. A specification curve shows how
estimates vary; the null-based criterion says which of those specifications deserve to be counted; the
permutation P value and the effective number say whether the share that survives is more than confounding
produces and how precisely it is known; and the calibrated decomposition says whether the analytic choice
that moves estimates most is evidence about the association or a property of the data structure. The
constraint test and the breakpoint diagnostics address two specific habits of the biomarker literature:
entering ratios without asking whether the ratio is the right combination, and reporting thresholds
whose location the data cannot fix.

## Reproducing the paper

`system.file("reproduce", package = "speccal")` lists scripts that rebuild the full 15,840-specification
grid, the 300-replicate null, the defensible subset, the variance decompositions, the 1,056 constraint
tests and the breakpoint bootstraps from the public NHANES files, and compare each with the archived
results ledger of the accompanying paper.
